Reconstructing the last known movements of one of Nature's giants

### 2018-02-01 manuscript/revision/blue-whale-ms-v2.tex
Typo

The blue whale (\textit{Balaenoptera musculus}), the largest animal to have ever lived, is one such migratory marine species. 
Despite hunting for blue whales being banned for over 50 years, blue whales are still listed as Endangered on the IUCN Red List\cite{reilly2008balaenoptera}.
In the northeast Atlantic, blue whales appear to form a separate population, with little evidence for exchange between northeast and northwest Atlantic populations\cite{pike2009note}.  
The northeast Atlantic population of blue whales was the first large whale population to be systematically hunted with explosive harpoons, and while the population was probably relatively small before hunting, now fewer than 1000 individuals are estimated to be in the northeast Atlantic in summer, mainly distributed around Iceland\cite{pike2009note}.
Following movements of individual blue whales is extremely challenging due to their scarcity and oceanic habit.
Similar to other large balaenopterid whales, blue whales in the northeast Atlantic are generally assumed to conduct annual migrations between high and low latitudes. 
A combination of historic whaling data, observations, satellite tracks and acoustic monitoring suggest that blue whales in the northeast Atlantic track regions of high production throughout the year, with at least some whales wintering in the upwelling systems between Mauritania and the Cape Verde Islands\cite{baines2014upwellings}.
Northward migration in the spring may occur along mid-Atlantic corridors with peak sightings in the Azores around April-May. Summer feeding appears to occur in northerly latitudes around Iceland, and historically in the Norwegian and Barents Seas\cite{pike2009note}.
Blue whales are frequently detected in waters to the west of the UK, with peak acoustic detections occurring between November and December, in southward-migrating animals\cite{reeves2004historical,baines2017autumn,charif2009acoustic,visser2011timing}.
While direct tagging methods are adding vital behavioural data to our understanding of whale movements\cite{borger15,mcdonald2006biogeographic,bailey2009behavioural,mate2007evolution,silva2013north}, they are often expensive, limited to short-term studies\cite{bailey2009behavioural,best2015tag,mate2007evolution}, and may influence whale behaviour\cite{walker2012review}.
Conservation efforts are consequently hampered by the lack of detailed data on the distribution, behaviour and ecology of blue whales, particularly during the winter and for the juvenile phase of life. 
An alternative technique for investigating whale movements retrospectively is to use intrinsic biochemical information such as stable isotopes\cite{west2006stable,busquets2017estimating,hobson2008tracking}. 
The stable isotope composition of animal tissues reflects the isotopic composition of diet at the time and place of ingestion, integrated over the timescale of tissue growth. 
Stable isotope data therefore provide information on diet and location (i.e. `trophic geography'\cite{bird2018global}), but incomplete knowledge of the spatio-temporal distribution of the isotopic composition of prey currently limits interpretation of measured values. 
Here we reconstruct movements of an historic whale by comparing isotopic compositions of samples from museum collections, rich archives of behavioural information\cite{lister2011natural}, to spatially referenced models of isotopic variation in diet across the northeast Atlantic \cite{hobson1999tracing,hobson2008tracking,eisenmann2016isotopic}.

>>>>

The blue whale (\textit{Balaenoptera musculus}), the largest animal to have ever lived, is one such migratory marine species. 
Despite hunting for blue whales being banned for over 50 years, blue whales are still listed as Endangered on the IUCN Red List\cite{reilly2008balaenoptera}.
In the northeast Atlantic, blue whales appear to form a separate population, with little evidence for exchange between northeast and northwest Atlantic populations\cite{pike2009note}.  
The northeast Atlantic population of blue whales was the first large whale population to be systematically hunted with explosive harpoons, and while the population was probably relatively small before hunting, now fewer than 1000 individuals are estimated to be in the northeast Atlantic in summer, mainly distributed around Iceland\cite{pike2009note}.
Following movements of individual blue whales is extremely challenging due to their scarcity and oceanic habit.
Similar to other large balaenopterid whales, blue whales in the northeast Atlantic are generally assumed to conduct annual migrations between high and low latitudes. 
A combination of historic whaling data, observations, satellite tracks and acoustic monitoring suggest that blue whales in the northeast Atlantic track regions of high production throughout the year, with at least some whales wintering in the upwelling systems between Mauritania and the Cape Verde islands\cite{baines2014upwellings}.
Northward migration in the spring may occur along mid-Atlantic corridors with peak sightings in the Azores around April-May. Summer feeding appears to occur in northerly latitudes around Iceland, and historically in the Norwegian and Barents Seas\cite{pike2009note}.
Blue whales are frequently detected in waters to the west of the UK, with peak acoustic detections occurring between November and December, in southward-migrating animals\cite{reeves2004historical,baines2017autumn,charif2009acoustic,visser2011timing}.
While direct tagging methods are adding vital behavioural data to our understanding of whale movements\cite{borger15,mcdonald2006biogeographic,bailey2009behavioural,mate2007evolution,silva2013north}, they are often expensive, limited to short-term studies\cite{bailey2009behavioural,best2015tag,mate2007evolution}, and may influence whale behaviour\cite{walker2012review}.
Conservation efforts are consequently hampered by the lack of detailed data on the distribution, behaviour and ecology of blue whales, particularly during the winter and for the juvenile phase of life. 
An alternative technique for investigating whale movements retrospectively is to use intrinsic biochemical information such as stable isotopes\cite{west2006stable,busquets2017estimating,hobson2008tracking}. 
The stable isotope composition of animal tissues reflects the isotopic composition of diet at the time and place of ingestion, integrated over the timescale of tissue growth. 
Stable isotope data therefore provide information on diet and location (i.e. `trophic geography'\cite{bird2018global}), but incomplete knowledge of the spatio-temporal distribution of the isotopic composition of prey currently limits interpretation of measured values. 
Here we reconstruct movements of an historic whale by comparing isotopic compositions of samples from museum collections, rich archives of behavioural information\cite{lister2011natural}, to spatially referenced models of isotopic variation in diet across the northeast Atlantic \cite{hobson1999tracing,hobson2008tracking,eisenmann2016isotopic}.
